---
title: "Bank Marketing Campaign EDA" subtitle: "What really drives term deposit
sign-ups?" author: "Sheriann McLarty" date: "`r format(Sys.Date(), '%B %d,
%Y')`" output: pdf_document: latex_engine: xelatex toc: true toc_depth: 3
html_document: toc: true toc_float: true number_sections: true df_print: paged
theme: journal fontsize: 11pt mainfont: Times New Roman geometry: margin=1in
header-includes:
  - \usepackage{graphicx}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{caption}
  - \captionsetup[table]{skip=6pt}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "85%"
)
# define show_code variable
show_code <- FALSE
```
```{r load libraries}
# Libraries
library(tidyverse)
library(corrplot)
library(kableExtra)
```
```{r load data}
# Data
bank <- read.csv("bank.csv", sep = ";")
# Helper: tidy prop table
prop_yes <- function(tbl){
  out <- as.data.frame(tbl)
  names(out) <- c("level","n")
  out$prop <- out$n / sum(out$n)
  out
}
# simple theme for kable
kbl_q <- function(df, caption = NULL, digits = 3){
  df %>%
    kbl(caption = caption, digits = digits, booktabs = TRUE) %>%
    kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped"))
}
```
> **Note:** This essay serves as the Executive Summary accompanying the
Exploratory Data Analysis (EDA) of the Bank Marketing Campaign dataset.
# Executive Summary & Findings
This dataset captures a Portuguese bank’s phone campaigns: 4,521 rows, 17
columns, a blend of numeric fields (age, balance, duration, campaign, pdays,
previous) and categorical fields (job, marital, education, contact, month,
poutcome). The target y indicates whether a client subscribed to a term deposit.
What matters most. The headline is imbalance and targeting. The target is skewed
about 89 percent “no” to 11 percent “yes.” If we do nothing, any model can “win”
by predicting “no” every time. Accuracy will look great and be useless. The real
lift comes from campaign design variables: previous outcome, contact method, and
timing, not simple demographics like age.
Central tendencies and spread. Age clusters in the late 30s to early 40s;
nothing wild there. Balance and duration are extremely right-skewed with long
tails. A handful of wealthy or very long-call cases will dominate a loss
function if left untreated. That means we should log-transform duration and
winsorize or log balance. Campaign count is concentrated around 1–3 calls;
behavior after that tends to diminish returns.
Categorical conversion rates (where the signal lives). When we compute
conversion rates by category, we see sharp differences:
Previous outcome (poutcome) is the strongest single indicator. A prior “success”
predicts a dramatically higher current conversion rate. This is textbook: past
positive engagement signals readiness.
Contact method matters: cell calls generally outperform unknown contact
channels.
Month matters: conversion rates tend to spike in a few months (often March and
December), while some high-volume months underperform. Timing is a lever.
Correlations are not the whole story. Numeric features have weak linear
correlation with the target; that is normal for marketing data. The heavy
lifting comes from categoricals and interactions (for example, job × month,
contact × previous outcome). That is exactly why linear correlation heatmaps
underwhelm here.
Missingness and duplicates. There are no formal NA values, but several fields
use "unknown". Treat those as explicit levels or flagged missing. Duplicates are
checked and removed if found (see Appendix code). This aligns with domain
reality: not every client discloses job/education, and many have never been
contacted before (pdays == -1).
Algorithm selection. This is a supervised binary classification problem. Use
Logistic Regression as a baseline for interpretability and stakeholder
communication (coefficients map to odds). Use Gradient Boosting (or Random
Forest) for performance, because it captures nonlinearities and interactions
that dominate marketing outcomes. If the dataset were under 1,000 rows, I would
prefer simpler models (Logistic/Decision Tree) to control overfitting.
Preprocessing playbook. Address class imbalance with SMOTE or class weights.
One-hot encode categoricals (job, education, contact, month, poutcome),
optionally keeping "unknown" as its own level. Log-transform duration; log or
winsorize balance; standardize numerics for linear models. Engineer high-value
flags such as contacted_before (pdays > 0) and seasonal targets (high-conversion
months). Evaluate with ROC-AUC and PR-AUC, not just accuracy.
Bottom line. To get more deposits, do not just dial more — dial smarter.
Prioritize clients with prior positive outcomes, use the best contact channel,
and time outreach in stronger months. Fix the imbalance, encode categoricals,
transform skew, and use Logistic Regression for storytelling with Gradient
Boosting for lift. That’s the play.
# Data Leakage Duration is the call length and is only known after the call. For
any pre-call scoring or campaign targeting, duration must be excluded from model
training and feature engineering. I will explicitly drop it in modeling chunks
meant for pre-call use. # Exploratory Data Analysis
**Structure and Shape**
```{r Structure and Shape}
dim(bank)
str(bank)
```
**Bottom line.** Mostly categorical features with a few skewed numerics. Target is y (yes/no).
## Preprocessing
```{r preprocessing}
## Preprocessing decisions reflected in later modeling
bank_prep <- bank %>%
  mutate(
    # pdays = -1 means "never contacted" — recode to a labeled factor
    pdays_cat = case_when(
      pdays == -1 ~ "never_contacted",
      pdays >= 0  ~ "contacted_before",
      TRUE        ~ "unknown"
    ),
    pdays_cat = factor(pdays_cat),
    # keep original for EDA; use *_cat in models
    # high-cardinality example handling (group rare levels)
    job_grp = fct_lump_n(factor(job), n = 10, other_level = "other_job"),
    education_grp = fct_lump_n(factor(education), n = 6, other_level = "other_edu")
  )
```
**Bottom-Line.** High-cardinality: I group infrequent job/education levels (top-N + “other”) to stabilize estimates and avoid sparse dummies. For tree models this is optional, but it simplifies interpretation and reduces variance.
## Evaluation Plan Classification threshold: With ~89% “no,” the default 0.50
cutoff is not optimal. I’ll tune the threshold on the validation set to maximize
F1 (or Youden’s J for ROC, or optimize Recall@Precision≥X for business needs).
Final reporting will include ROC-AUC and PR-AUC (better reflects minority-class
lift) along with a threshold-tuned confusion matrix.
## Conistency Checks + "unknown" + duplicates

```{r data checks}
## Data Consistency & Integrity
# 1) exact duplicate rows
dup_rows <- sum(duplicated(bank))

# 2) structural NA vs sentinel "unknown"
unknown_share <- bank %>%
  summarise(across(everything(), ~mean(. == "unknown")*100)) %>%
  pivot_longer(everything(), names_to="variable", values_to="pct_unknown") %>%
  arrange(desc(pct_unknown))
# 3) impossible ranges sanity checks (examples)
range_checks <- tibble(
  check = c("age >= 18", "campaign >= 1", "pdays == -1 allowed"),
  valid = c(all(bank$age >= 18), all(bank$campaign >= 1), any(bank$pdays == -1))
)
kbl_q(tibble(Metric=c("Duplicate rows"), Value=c(dup_rows)), "Data Quality Checks")
kbl_q(unknown_share, "Share of 'unknown' by Variable (%)", digits = 1)
kbl_q(range_checks, "Range/Sentinel Sanity Checks")
```
## Central Tendency & Spread
```{r Central Tendency & Spread}
num_summary <- bank %>%
  select(age, balance, duration, campaign, pdays, previous) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  group_by(variable) %>%
  summarise(
    n      = n(),
    mean   = mean(value, na.rm = TRUE),
    sd     = sd(value, na.rm = TRUE),
    min    = min(value, na.rm = TRUE),
    q25    = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q75    = quantile(value, 0.75, na.rm = TRUE),
    max    = max(value, na.rm = TRUE),
    .groups = "drop"
  )

kbl_q(num_summary, "Summary of Numeric Predictors")
```
**Why it matters.** Balance and duration are heavily right-skewed. Outliers will hijack training unless we log/trim.
## Distributions
**The Shape Of The Problem**
```{r distrubution and shape}
bank %>%
  select(age, balance, duration, campaign) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  facet_wrap(~name, scales = "free", nrow = 2) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Key Numeric Distributions")
```
**Bottom line.** Transform balance and duration; treat campaign’s heavy mass at 1–3 calls carefully.
## Target Balance
**Imbalance Check**
```{r imbalane check}
prop_yes(table(bank$y)) %>%
  mutate(level = factor(level, levels = c("no","yes"))) %>%
  arrange(level) %>%
  kbl_q("Target Class Balance", digits = 3)
```
**Bottom line.** With about 89 percent “no,” accuracy alone is a trap. Optimize recall/precision on “yes.”
## Correlations **Numeric Only**
```{r correlations}
num_vars <- bank %>% select(where(is.numeric))
cor_mat  <- cor(num_vars, use = "pairwise.complete.obs")
corrplot(cor_mat, method = "color", type = "upper", tl.cex = 0.8)
```
**Bottom line**. Weak linear correlations are expected; the real signal is in categoricals and interactions.
## Categorical Conversion
**Where The Signal Lives**
```{r cat conversion}
conv_job <- bank %>%
  group_by(job, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = ifelse(total > 0, yes/total, NA_real_)) %>%
  arrange(desc(conversion_rate))
conv_job %>%
  select(job, yes, no, total, conversion_rate) %>%
  head(8) %>%
  kbl_q("Top Job Categories by Conversion Rate", digits = 3)
conv_contact <- bank %>%
  group_by(contact, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = yes/total) %>%
  arrange(desc(conversion_rate))
conv_poutcome <- bank %>%
  group_by(poutcome, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = yes/total) %>%
  arrange(desc(conversion_rate))
kbl_q(conv_contact %>% select(contact, yes, no, total, conversion_rate),
      "Contact Method (Conversion Rate)", digits = 3)
kbl_q(conv_poutcome %>% select(poutcome, yes, no, total, conversion_rate),
      "Previous Outcome (Conversion Rate)", digits = 3)
conv_month <- bank %>%
  group_by(month, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = yes/total) %>%
  arrange(desc(conversion_rate))
kbl_q(conv_month, "Conversion by Month", digits = 3)
```
**Bottom line.** Prior positive outcome, contact method, and timing are the levers. That is your targeting strategy.
## Missingness *The "Unknown" Game*
```{r missingness}
missing_tbl <- bank %>%
  summarise(across(everything(), ~mean(. == "unknown")*100)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_unknown") %>%
  arrange(desc(pct_unknown))
kbl_q(missing_tbl, "Share of 'unknown' by Variable (%)", digits = 1)
```
**Bottom line.** Keep "unknown" as an explicit level or flag it. Do not silently drop it.
# Outliers *Trim Or Transform?*
```{r outliers}
bank %>%
  select(balance, duration, campaign) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = name, y = value)) +
  geom_boxplot(fill = "orange") +
  coord_cartesian(ylim = c(-1000, 5000)) +
  labs(title = "Outliers in Key Numerics (Trimmed View)")
```
**Bottom line.** Log duration; log or winsorize balance; be cautious beyond 95th–99th percentiles.
## Duplicates & Consistency Check
```{r duplicate and consistency check}
dup_n <- sum(duplicated(bank))
kbl_q(tibble(Metric = c("Duplicate rows"), Value = dup_n), "Data Quality Check", digits = 0)
```
*Bottom line.** Remove duplicates if any appear; keep one distinct record.
# Modeling Strategy *What Fits The Business*
Logistic Regression — interpretable coefficients for stakeholders; great
baseline; needs one-hot encoding and scaling.
Gradient Boosting / Random Forest — stronger accuracy, handles nonlinearities
and interactions; less transparent; use SHAP for explainability.
Small data (< 1000 rows) — prefer simpler models (Logistic/Decision Tree) to
avoid overfitting.
*Recommendation:* baseline with Logistic Regression for storytelling; deploy Gradient Boosting for lift. Evaluate with ROC-AUC and PR-AUC, not plain accuracy.
## Preprocessing Plan *Execution Playbook*
Imbalance: SMOTE or class weights; report ROC-AUC and PR-AUC.
Categoricals: one-hot encode job, education, contact, month, poutcome; keep
"unknown" explicit or add missing flags.
Numerics: log-transform duration; log or winsorize balance; standardize for
linear models.
Features: contacted_before (pdays > 0); seasonal flags for high-conversion
months; optional interactions.
Sanity: drop duplicates; trim extreme values where appropriate.
# Conclusion
If the goal is more deposits, do not just dial more — dial smarter. Prioritize
people with prior positive outcomes, use the right contact method, and time the
calls when conversion is historically stronger. Fix the imbalance, encode
categoricals, transform skewed numerics, and pair Logistic Regression
(explainability) with Gradient Boosting (performance). That is the plan I would
ship.
\newpage \addtocontents{toc}{\protect\setcounter{tocdepth}{0}} # Appendix: R
Full Code
```{r,full code, echo=TRUE}
# Libraries
library(tidyverse)
library(corrplot)
library(kableExtra)
# Data
bank <- read.csv("bank.csv", sep = ";")
# Dimensions & structure
dim(bank); str(bank)
## Preprocessing decisions reflected in later modeling
bank_prep <- bank %>%
  mutate(
    # pdays = -1 means "never contacted" — recode to a labeled factor
    pdays_cat = case_when(
      pdays == -1 ~ "never_contacted",
      pdays >= 0  ~ "contacted_before",
      TRUE        ~ "unknown"
    ),
    pdays_cat = factor(pdays_cat),
    # keep original for EDA; use *_cat in models
    # high-cardinality example handling (group rare levels)
    job_grp = fct_lump_n(factor(job), n = 10, other_level = "other_job"),
    education_grp = fct_lump_n(factor(education), n = 6, other_level = "other_edu")
  )
## Data Consistency & Integrity
# 1) exact duplicate rows
dup_rows <- sum(duplicated(bank))
# 2) structural NA vs sentinel "unknown"
unknown_share <- bank %>%
  summarise(across(everything(), ~mean(. == "unknown")*100)) %>%
  pivot_longer(everything(), names_to="variable", values_to="pct_unknown") %>%
  arrange(desc(pct_unknown))
# 3) impossible ranges sanity checks (examples)
range_checks <- tibble(
  check = c("age >= 18", "campaign >= 1", "pdays == -1 allowed"),
  valid = c(all(bank$age >= 18), all(bank$campaign >= 1), any(bank$pdays == -1))
)
kbl_q(tibble(Metric=c("Duplicate rows"), Value=c(dup_rows)), "Data Quality Checks")
kbl_q(unknown_share, "Share of 'unknown' by Variable (%)", digits = 1)
kbl_q(range_checks, "Range/Sentinel Sanity Checks")
# Central tendency
summary(select(bank, age, balance, duration, campaign, pdays, previous))
# Distributions
bank %>%
  select(age, balance, duration, campaign) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  facet_wrap(~name, scales = "free", nrow = 2) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Key Numeric Distributions")
# Target balance
as.data.frame(prop.table(table(bank$y)))
# Correlations (numeric only)
num_vars <- bank %>% select(where(is.numeric))
cor_mat  <- cor(num_vars, use = "pairwise.complete.obs")
corrplot(cor_mat, method = "color", type = "upper", tl.cex = 0.8)
# Categorical conversion rates
conv_job <- bank %>%
  group_by(job, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = yes / total) %>%
  arrange(desc(conversion_rate))
head(conv_job, 10)
conv_contact <- bank %>%
  group_by(contact, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = yes / total) %>%
  arrange(desc(conversion_rate))
conv_contact
conv_poutcome <- bank %>%
  group_by(poutcome, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = yes / total) %>%
  arrange(desc(conversion_rate))
conv_poutcome
conv_month <- bank %>%
  group_by(month, y) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = y, values_from = n, values_fill = 0) %>%
  rename(no = `no`, yes = `yes`) %>%
  mutate(total = no + yes, conversion_rate = yes / total) %>%
  arrange(desc(conversion_rate))
conv_month
# Missingness as % "unknown"
missing_tbl <- bank %>%
  summarise(across(everything(), ~mean(. == "unknown")*100)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_unknown") %>%
  arrange(desc(pct_unknown))
missing_tbl
# Outliers (trimmed view)
bank %>%
  select(balance, duration, campaign) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = name, y = value)) +
  geom_boxplot(fill = "orange") +
  coord_cartesian(ylim = c(-1000, 5000)) +
  labs(title = "Outliers in Key Numerics (Trimmed)")
# Duplicate check
sum(duplicated(bank))
```
\addtocontents{toc}{\protect\setcounter{tocdepth}{2}}}